Equivalence Hypergraphs:\\ Semantics of E-Graphs and their Generalization to Monoidal Theories

### 2023-10-09 sections/combinatorial_semantics.tex
some clean-up

@@

% TODO: fix union to Kleene star, introduce notation, put explanation out of def environmentnt, fix notation for in(e) and out(e), conflict relation is not transitive

@@

Before we begin we first fix some notation. When $V$ is a set, we denote with $V^{*}$ a set of all finite sequences of the elements of $V$, i.e. $^*$ is a Kleene star. When $\phi : V \to V'$ is a function, we denote with $\phi^{*}$ an extension of $\phi$ to sequences, i.e. $\phi^{*} : V^* \to V'^{*}$ and it applies $\phi$ element-wise. Similarly, in the case when $\psi \subset V \times V$ is a relation: we denote with $\psi^{*}: \subset V^* \times V^*$ the element-wise relation on ordered sequences.

@@

\begin{definition}[Hypergraph]
\label{def:hypergraph}

A hypergraph $\mathcal{G}_{\Sigma}$ over a monoidal signature $\Sigma$ is a tuple 
\[
(V_{\mathcal{G}},E_{\mathcal{G}},s,t,l)
\]

@@

We will omit the subscripts when they are clear from the context.
\end{definition}

\begin{definition}[Hypergraph homomorphism]

@@

where $\phi^{*}$ is an element-wise application of $\phi$.
\end{definition}

@@

A morphism $\alpha : (X \xrightarrow{} A \xleftarrow{} Y) \Rightarrow (X \xrightarrow{} B \xleftarrow{} Y)$ between cospans is an arrow $\alpha : A \to B$ in $\mathbb{C}$ such that the following diagram commutes.

@@

Cospans are isomorphic if $\alpha$ is an isomorphism. For $X \in \mathbb{C}$ an $id$-cospan is $X \xrightarrow{id_X} X \xleftarrow{id_X} X$. The composition of cospans $X \xrightarrow{} A \xleftarrow{f} Y$ and $Y \xrightarrow{g} B \xleftarrow{} Z$ is $X \xrightarrow{} A +_{f,g} B \xleftarrow{} Z$, constructed by taking a pushout of $f,g$. This data is the category $\catname{Csp(\mathbb{C})}$: the objects are those of $\mathbb{C}$ and the arrows are isomorphism classes of cospans. Finally, $\catname{Csp(\mathbb{C})}$ is a monoidal category: monoidal product is given by a coproduct (denoted with $+$) in $\mathbb{C}$ and the unit is given by the initial object of $\mathcal{C}$ as can be seen below.

\[
\adjustbox{scale=0.75,center}{
% https://q.uiver.app/#q=WzAsMTQsWzAsMCwibiJdLFsxLDAsIlxcbWF0aGNhbHtHfSJdLFsyLDAsIm0iXSxbMSwxLCJcXG90aW1lcyJdLFswLDIsIm4nIl0sWzIsMiwibSciXSxbMSwyLCJcXG1hdGhjYWx7Ryd9Il0sWzMsMSwiOj0iXSxbNCwxLCJuK24nIl0sWzYsMSwiXFxtYXRoY2Fse0d9K1xcbWF0aGNhbHtHJ30iXSxbOCwxLCJtK20nIl0sWzEwLDEsIjAiXSxbMTEsMSwiMCJdLFsxMiwxLCIwIl0sWzAsMSwiZiJdLFsyLDEsImciLDJdLFs0LDYsImYnIl0sWzUsNiwiZyciLDJdLFs4LDksImYrZiciXSxbMTAsOSwiZytnJyIsMl0sWzExLDEyLCJpZF8wIl0sWzEzLDEyLCJpZF8wIiwyXV0=
\begin{tikzcd}
	n & {\mathcal{G}} & m \\
	& \otimes && {:=} & {n+n'} && {\mathcal{G}+\mathcal{G'}} && {m+m'} && 0 & 0 & 0 \\
	{n'} & {\mathcal{G'}} & {m'}
	\arrow["f", from=1-1, to=1-2]
	\arrow["g"', from=1-3, to=1-2]
	\arrow["{f'}", from=3-1, to=3-2]
	\arrow["{g'}"', from=3-3, to=3-2]
	\arrow["{f+f'}", from=2-5, to=2-7]
	\arrow["{g+g'}"', from=2-9, to=2-7]
	\arrow["{id_0}", from=2-11, to=2-12]
	\arrow["{id_0}"', from=2-13, to=2-12]
\end{tikzcd}
}
\]

Symmetry is inherited from $\mathcal{C}$ as well as the coproduct is commutative and is given below

@@

When $\mathbb{C} = Hyp_{\Sigma}$ we get a category of cospans of hypergraphs. An empty hypergraph is the initial object in $Hyp_{\Sigma}$ and coproduct and pushout are defined on the underlying sets of nodes and edges.
Following~\cite{Frobenius2} morphisms in an SMC with a signature $\Sigma$ can be interpreted as particular cospans of hypergraphs.

\begin{definition}[Category $\;\catname{Csp_{D}(Hyp_{\Sigma})}$]
\label{def:cspd}
Category $\catname{Csp_{D}(Hyp_{\Sigma})}$ is a sub-category of $\catname{Csp(Hyp_{\Sigma})}$ which has discrete hypergraphs as objects.
\end{definition}

In particular, this means that any morphism in $\catname{Csp_{D}(Hyp_{\Sigma})}$ is of the form

\[
n \xrightarrow{} \mathcal{F} \xleftarrow{} m
\]

where $n,\;m$ are discrete hypergraphs and $\mathcal{F}$ is an arbitrary hypergraph. When string diagrams are interpreted as cospans of hypergraphs $n,m$ correspond to input and output interfaces of a string diagram respectively.

A pictorial representation of such a cospan is depicted below in Figure~\ref{fig:hypergraph_and_string_diagram}. We use node labels to define mappings $n \to \mathcal{F}$ and $m \to \mathcal{F}$.

\begin{figure}
    \centering
    \begin{subfigure}{0.45\linewidth}
    \[
    \scalebox{0.75}{
	\tikzfig{combinatorial_semantics/mda-hypergraph-example}
   }
   \]
    \end{subfigure}
    \begin{subfigure}{0.45\linewidth}
        \[
        \scalebox{0.75}{
        \tikzfig{combinatorial_semantics/mda-hypergraph-example-string}
        }
        \]
    \end{subfigure}
    \caption{Cospan of hypergraphs and corresponding string diagram}
    \label{fig:hypergraph_and_string_diagram}
\end{figure}

@@

\begin{definition}[Path]
    A path from $e_0$ to $e_{n-1}$ of length $n$ in a hypergraph $\mathcal{F}$ is an ordered sequence of edges $e_i \in E_{F}$ : $[e_0, e_1, \ldots, e_{n-1}]$ such that for all $i < n - 1$ some node $v \in t(e_i)$ is also in $s(e_{i+1})$.
\end{definition}

@@

We call a cospan $n \xrightarrow{f} \mathcal{F} \xleftarrow{g} m$ monogamous directed acyclic if

\begin{enumerate}
    \item $\mathcal{F}$ is a directed acyclic hypergraph
    \item $f$ and $g$ are monos
    \item the in-degree and out-degree of every vertex of $\mathcal{F}$ is at most 1
    \item vertices of $\mathcal{F}$ with in-degree 0 are precisely the image of $f$
    \item vertices of $\mathcal{F}$ with out-degree 0 are precisely the image of $g$
\end{enumerate}

@@

\begin{definition}[Category $\catname{MACsp_{D}(Hyp_{\Sigma})}$]
    When we narrow our attention to cospans (morphisms) of $\catname{Csp_{D}(Hyp_{\Sigma})}$ which are monogamous we get the category of monogamous directed acyclic (mda-) cospans of hypegraphs $\catname{MACsp_{D}(Hyp_{\Sigma})}$. Vertical and horizontal compositions of mda-cospans are mda-cospans, identities and symmetries are mda-cospans, and hence this is a proper category~\cite{Frobenius2}.
\end{definition}

\begin{theorem}[Corollary 26~\cite{Frobenius2}]
    $\textsf{PROP}(\Sigma) \cong \catname{MACsp_{D}(Hyp_{\Sigma})}$
\end{theorem}
This theorem essentially states a correspondence between morphisms in SMC (string diagrams) and mda-cospans of hypergraphs.
% This section presents a combinatorial structure that defines our e-hypergraphs.

% We closely follow a set of definitions designed in~\cite{fscd}. We extend this set to support an additional e-structure.

@@

@@

In other words, homomorphism should preserve immediate predecessors and preserve and reflect the conflict relation.

@@

\begin{definition}[Category $\catname{EHyp_{\Sigma}}$]
Such e-hypergraphs form a category $\catname{EHyp_{\Sigma}}$ where morphisms are e-hypergraph homomorphisms defined above.
\end{definition}

\begin{remark}
    Similarly to $\catname{Hyp_{\Sigma}}$ pushouts in $\catname{EHyp_{\Sigma}}$ are computed on the underlying sets of nodes and edges. However, unlike $\catname{Hyp_{\Sigma}}$ $\catname{EHyp_{\Sigma}}$ is not \textit{adhesive} meaning that not all pushouts along monos exist. Consider an example span below in Figure~\ref{fig:pushout_fail}.

\begin{figure}[h!]
    \centering
    \[    
    \scalebox{0.5}{\tikzfig{combinatorial_semantics/pushout_fail_example}
    }
    \]
    \caption{Pushout failure}
    \label{fig:pushout_fail}
\end{figure}

    In order to build a commuting square which is a pushout the immediate predecessors of $f$ should be identified in the pushout because any element of an e-hypergraph has only one immediate predecessor. However, such identification is not possible as these predecessors have different types. Luckily, we do not need arbitrary pushouts and pushouts that we actually need do exist.
\end{remark}

@@

\update{I will build below $\catname{ECsp_{D}(EHyp_{\Sigma})}$ (E for extended) instead of $ECsp(\mathcal{C})$ for arbitrary $\mathcal{C}$ with colimits to avoid talking about pushouts in general below, because we use set difference.}
\begin{definition}[Category $\catname{ECsp_{D}(EHyp_{\Sigma})}$]
    We will build the category of \textit{extended} cospans of e-hypergraphs as follows.
    Let the objects of $\catname{ECsp_{D}(EHyp_{\Sigma})}$ be only discrete e-hypergraphs of $\catname{EHyp_{\Sigma}}$ and morphisms from $ext(X) \to ext(Y)$ be the cospans (ext stands for external)

    \[
    ext(X) \xrightarrow{f_{ext}} X \xrightarrow{f} A \xleftarrow{g} Y \xleftarrow{g_{ext}} ext(Y)
    \] where $A \in obj(EHyp_{\Sigma})$.

    \update{Can I say that a morphism between two such cospans is 3 arrows?}

@@

@@

\[
\begin{tikzcd}
	& X & A & {~Y} \\
	{ext(X)} &&&& {ext(Y)} \\
	& {X'} & {A'} & {Y'}
	\arrow[curve={height=-12pt}, from=2-1, to=1-2]
	\arrow[from=1-2, to=1-3]
	\arrow[from=1-4, to=1-3]
	\arrow[curve={height=12pt}, from=2-5, to=1-4]
	\arrow[curve={height=-12pt}, from=2-5, to=3-4]
	\arrow[from=3-4, to=3-3]
	\arrow[from=3-2, to=3-3]
	\arrow[curve={height=12pt}, from=2-1, to=3-2]
	\arrow["\alpha"{description}, from=1-2, to=3-2]
	\arrow["\beta"{description}, from=1-3, to=3-3]
	\arrow["\gamma"{description}, from=1-4, to=3-4]
\end{tikzcd}
\]

@@

Composition of two morphisms $f : ext(X) \to ext(Y)$ and $g : ext(Y) \to ext(Z)$ is computed in two stages. First, $A +_{f_{ext};f, g_{ext};g} B$ is computed which is a pushout below

% https://q.uiver.app/#q=WzAsMTAsWzAsMCwiZXh0KFgpIl0sWzEsMSwiWCJdLFsyLDIsIkEiXSxbMywwLCJZIl0sWzQsMCwiZXh0KFkpIl0sWzUsMCwiWSciXSxbNiwyLCJCIl0sWzcsMSwiWiJdLFs4LDAsImV4dChaKSJdLFs0LDMsIl57XFx1bGNvcm5lcn1BK197Zl97ZXh0fTtmLGdfe2V4dH07Z31CIl0sWzAsMV0sWzEsMl0sWzMsMiwiZiJdLFs0LDMsImZfe2V4dH0iXSxbNCw1LCJnX3tleHR9IiwyXSxbNSw2LCJnIiwyXSxbNyw2XSxbOCw3XSxbMiw5XSxbNiw5XV0=
\[\begin{tikzcd}
	{ext(X)} &&& Y & {ext(Y)} & {Y'} &&& {ext(Z)} \\
	& X &&&&&& Z \\
	&& A &&&& B \\
	&&&& {^{\ulcorner}A+_{f_{ext};f,g_{ext};g}B}
	\arrow[from=1-1, to=2-2]
	\arrow[from=2-2, to=3-3]
	\arrow["f", from=1-4, to=3-3]
	\arrow["{f_{ext}}", from=1-5, to=1-4]
	\arrow["{g_{ext}}"', from=1-5, to=1-6]
	\arrow["g"', from=1-6, to=3-7]
	\arrow[from=2-8, to=3-7]
	\arrow[from=1-9, to=2-8]
	\arrow[from=3-3, to=4-5]
	\arrow[from=3-7, to=4-5]
\end{tikzcd}
\]

Then, new legs for $A +_{f_{ext};f,g_{ext};g} B$ are computed as below

\[
ext(X) \xrightarrow{f_{ext}} X \uplus (Y' \setminus ext(Y)) \xrightarrow{h} A +_{f_{ext};f,g_{ext};g} B \xleftarrow{k} Z \uplus (Y \setminus ext(Y)) \xleftarrow{g_{ext}} ext(Z)
\]

since $X,Y,Z,Y', ext(Y)$ are all discrete, $\uplus$ and $\setminus$ are standard set operations. $\setminus$ could be also formalised further as the following pushout complement

\[
\begin{tikzcd}
ext(Y) \arrow[d] & 0 \arrow[l] \arrow[d]    \\
Y'          & Y' \setminus ext(Y) \arrow[l]
\end{tikzcd}
\]

and then the disjoint union is just a coproduct. Defining cospans up to isomorphism, associativity of coproducts and pushouts, and the presence of an id-cospan turns this construction into a valid category.

Similarly to~\ref{def:cspd} this category has a symmetric monoidal product given by a coproduct in $\catname{EHyp_{\Sigma}}$ which makes it into a SMC. More precisely, below are cospans for tensor product, its unit and symmetry.

% https://q.uiver.app/#q=WzAsMTcsWzAsMCwiZXh0KFgpIl0sWzEsMCwiWCJdLFsyLDAsIkEiXSxbMywwLCJZIl0sWzQsMCwiZXh0KFkpIl0sWzAsMiwiZXh0KFgnKSJdLFsxLDIsIlgnIl0sWzIsMiwiQiJdLFszLDIsIlknIl0sWzQsMiwiZXh0KFknKSJdLFsyLDEsIlxcb3RpbWVzIl0sWzUsMSwiOj0iXSxbNiwxLCJleHQoWCkrZXh0KFgnKSJdLFs4LDEsIlgrWCciXSxbMTAsMSwiQStCIl0sWzEyLDEsIlkrWSciXSxbMTQsMSwiZXh0KFkpK2V4dChZJykiXSxbMCwxLCJmX3tleHR9Il0sWzQsMywiZ197ZXh0fSIsMl0sWzMsMiwiZyIsMl0sWzEsMiwiZiJdLFs1LDYsImYnX3tleHR9Il0sWzYsNywiZiciXSxbOCw3LCJnJyIsMl0sWzksOCwiZydfe2V4dH0iLDJdLFsxMiwxMywiZl97ZXh0fStmJ197ZXh0fSJdLFsxMywxNCwiZitmJyJdLFsxNSwxNCwiZytnJyIsMl0sWzE2LDE1LCJnX3tleHR9ICsgZydfe2V4dH0iLDJdXQ==
\[
\adjustbox{scale=0.6, center}{
\begin{tikzcd}
	{ext(X)} & X & A & Y & {ext(Y)} \\
	&& \otimes &&& {:=} & {ext(X)+ext(X')} && {X+X'} && {A+B} && {Y+Y'} && {ext(Y)+ext(Y')} \\
	{ext(X')} & {X'} & B & {Y'} & {ext(Y')}
	\arrow["{f_{ext}}", from=1-1, to=1-2]
	\arrow["{g_{ext}}"', from=1-5, to=1-4]
	\arrow["g"', from=1-4, to=1-3]
	\arrow["f", from=1-2, to=1-3]
	\arrow["{f'_{ext}}", from=3-1, to=3-2]
	\arrow["{f'}", from=3-2, to=3-3]
	\arrow["{g'}"', from=3-4, to=3-3]
	\arrow["{g'_{ext}}"', from=3-5, to=3-4]
	\arrow["{f_{ext}+f'_{ext}}", from=2-7, to=2-9]
	\arrow["{f+f'}", from=2-9, to=2-11]
	\arrow["{g+g'}"', from=2-13, to=2-11]
	\arrow["{g_{ext} + g'_{ext}}"', from=2-15, to=2-13]
\end{tikzcd}
}
\]

% https://q.uiver.app/#q=WzAsNSxbMCwwLCIwIl0sWzEsMCwiMCJdLFsyLDAsIjAiXSxbMywwLCIwIl0sWzQsMCwiMCJdLFswLDEsImlkXzAiXSxbMSwyLCJpZF8wIl0sWzMsMiwiaWRfMCIsMl0sWzQsMywiaWRfMCIsMl1d
\[
\adjustbox{scale=0.75, center}{
\begin{tikzcd}
	0 & 0 & 0 & 0 & 0
	\arrow["{id_0}", from=1-1, to=1-2]
	\arrow["{id_0}", from=1-2, to=1-3]
	\arrow["{id_0}"', from=1-4, to=1-3]
	\arrow["{id_0}"', from=1-5, to=1-4]
\end{tikzcd}
}
\]

@@

\begin{remark}
    When writing things like $Y' \setminus ext(Y)$ we mean $Y' \setminus g(Y')$ and so on.
\end{remark}

>>>>

% !TEX root = ../main.tex

@@

@@

Before we begin we first fix some notation.
When $V$ is a set, we denote with $V^{*}$ a set of all finite sequences of the elements of $V$, i.e. $^*$ is a free monoid.
When $\phi : V \to V'$ is a function, we denote with $\phi^{*}$ an extension of $\phi$ to sequences, i.e. $\phi^{*} : V^* \to V'^{*}$ and it applies $\phi$ element-wise.
Similarly, in the case when $\psi \subset V \times V$ is a relation: we denote with $\psi^{*}: \subset V^* \times V^*$ the element-wise relation on ordered sequences.
We will denote with $A +_{f,g} B$ the pushout of $A \xleftarrow{f} C \xrightarrow{g} B$.

@@

A hypergraph $\mathcal{G}_{\Sigma}$ over a monoidal signature $\Sigma$ is a tuple 
\[
(V_{\mathcal{G}},E_{\mathcal{G}},s,t,l)
\]
\label{def:hypergraph}

@@

We will omit the subscripts when they are clear from the context.

@@

@@

Two cospans $X \xrightarrow{} A \xleftarrow{} Y$ and $X \xrightarrow{} B \xleftarrow{} Y$ are isomorphic if the following diagram commutes and $\alpha$ is iso.

@@

For $X \in \mathbb{C}$ an $id$-cospan is $X \xrightarrow{id_X} X \xleftarrow{id_X} X$.
The composition of cospans $X \xrightarrow{} A \xleftarrow{f} Y$ and $Y \xrightarrow{g} B \xleftarrow{} Z$ is $X \xrightarrow{} A +_{f,g} B \xleftarrow{} Z$, constructed by taking a pushout of $f,g$.
This data is the category $\catname{Csp(\mathbb{C})}$: the objects are those of $\mathbb{C}$ and the arrows are isomorphism classes of cospans.
Finally, $\catname{Csp(\mathbb{C})}$ is a monoidal category: monoidal product is given by a coproduct (denoted with $+$) in $\mathbb{C}$ and the unit is given by the initial object of $\mathcal{C}$ as can be seen below.

% \[
% \adjustbox{scale=0.75,center}{
% https://q.uiver.app/#q=WzAsMTEsWzAsMCwiQSJdLFsxLDAsIlxcbWF0aGNhbHtHfSJdLFsyLDAsIkIiXSxbMywwLCJcXG90aW1lcyJdLFs0LDAsIkEnIl0sWzYsMCwiQiciXSxbNSwwLCJcXG1hdGhjYWx7Ryd9Il0sWzcsMCwiOj0iXSxbOCwwLCJBK0EnIl0sWzEwLDAsIlxcbWF0aGNhbHtHfStcXG1hdGhjYWx7Ryd9Il0sWzEyLDAsIkIrQiciXSxbMCwxLCJmIl0sWzIsMSwiZyIsMl0sWzQsNiwiZiciXSxbNSw2LCJnJyIsMl0sWzgsOSwiZitmJyJdLFsxMCw5LCJnK2cnIiwyXV0=
% \begin{tikzcd}
% 	A & {\mathcal{G}} & B & \otimes & {A'} & {\mathcal{G'}} & {B'} & {:=} & {A+A'} && {\mathcal{G}+\mathcal{G'}} && {B+B'}
% 	\arrow["f", from=1-1, to=1-2]
% 	\arrow["g"', from=1-3, to=1-2]
% 	\arrow["{f'}", from=1-5, to=1-6]
% 	\arrow["{g'}"', from=1-7, to=1-6]
% 	\arrow["{f+f'}", from=1-9, to=1-11]
% 	\arrow["{g+g'}"', from=1-13, to=1-11]
% \end{tikzcd}
\[
    A \xrightarrow{f} \mathcal{G} \xleftarrow{g} B \otimes A' \xrightarrow{f'} \mathcal{G'} \xleftarrow{g'} B' := A + A' \xrightarrow{f'} \mathcal{G} + \mathcal{G'} \xleftarrow{g'} B + B'
\]

Symmetry is inherited from $\mathcal{C}$ as the coproduct is commutative and is given below.

@@

When $\mathbb{C} = \catname{Hyp_{\Sigma}}$ we get a category of cospans of hypergraphs.
$\catname{Hyp_{\Sigma}}$ has all finite colimits and, in particular, coproduct is a disjoint union of two hypergraphs.
An empty hypergraph is the initial object in $Hyp_{\Sigma}$ and coproduct and pushout are defined on the underlying sets of nodes and edges.
Following~\cite{Frobenius2} morphisms in an SMC with a signature $\Sigma$ can be interpreted as particular cospans of hypergraphs.

\begin{definition}[Category $\;\catname{Csp_{D}(Hyp_{\Sigma})}$]
\label{def:cspd}
Category $\catname{Csp_{D}(Hyp_{\Sigma})}$ is a \textit{full} sub-category of $\catname{Csp(Hyp_{\Sigma})}$ which has discrete hypergraphs as objects.
\end{definition}

In particular, this means that any morphism in $\catname{Csp_{D}(Hyp_{\Sigma})}$ is of the form $n \xrightarrow{} \mathcal{G} \xleftarrow{} m$,
where $n,\;m$ are discrete hypergraphs and $\mathcal{G}$ is an arbitrary hypergraph.
When string diagrams are interpreted as cospans of hypergraphs $n,m$ correspond to input and output interfaces of a string diagram respectively.
A pictorial representation of such a cospan is depicted below in Figure~\ref{fig:hypergraph_and_string_diagram}.
We use node labels to define mappings $n \to \mathcal{F}$ and $m \to \mathcal{F}$.

\begin{figure}
    \centering
    \begin{subfigure}{0.45\linewidth}
    \[
    \scalebox{0.6}{
	\tikzfig{combinatorial_semantics/mda-hypergraph-example}
   }
   \]
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\linewidth}
        \[
        \raisebox{-0.6cm}{
        \scalebox{0.75}{
        \tikzfig{combinatorial_semantics/mda-hypergraph-example-string}
        }}
        \]
    \end{subfigure}
    \caption{Cospan of hypergraphs and corresponding string diagram}
    \label{fig:hypergraph_and_string_diagram}
\end{figure}

@@

\begin{definition}[Path]
    A path from $e_0$ to $e_{n-1}$ of length $n$ in a hypergraph $\mathcal{G}$ is an ordered sequence of edges $e_i \in E_{G}$ : $[e_0, e_1, \ldots, e_{n-1}]$ such that for all $i < n - 1$ some node $v \in t(e_i)$ is also in $s(e_{i+1})$.
\end{definition}

@@

We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\catname{Csp_{D}(Hyp_{\Sigma})}$ monogamous directed acyclic if

\begin{enumerate}
    \item $\mathcal{G}$ is a directed acyclic hypergraph
    \item $f$ and $g$ are monos
    \item the in-degree and out-degree of every vertex of $\mathcal{G}$ is at most 1
    \item vertices of $\mathcal{G}$ with in-degree 0 are precisely the image of $f$
    \item vertices of $\mathcal{G}$ with out-degree 0 are precisely the image of $g$
\end{enumerate}

@@

When we narrow our attention to cospans (morphisms) of $\catname{Csp_{D}(Hyp_{\Sigma})}$ which are monogamous we get the category of monogamous directed acyclic (mda-) cospans of hypegraphs $\MdaCospans$.

\begin{definition}[Category $\MdaCospans$]
    We define $\MdaCospans$ as a subcategory of $\catname{Csp_{D}(Hyp_{\Sigma})}$.
\end{definition}

Vertical and horizontal compositions of mda-cospans are mda-cospans, identities and symmetries are mda-cospans, tensor unit is discrete, and hence this is a proper symmetric monoidal category~\cite{Frobenius2}.

\begin{theorem}[Corollary 26~\cite{Frobenius2}]
    $\textsf{PROP}(\Sigma) \cong \MdaCospans$
\end{theorem}
This theorem essentially states a correspondence between morphisms in SMC (string diagrams) and mda-cospans of hypergraphs.

@@

\Aleksei{Switch to the complent in the definition of the relation and call it a \textit{consistency} relation}

@@

In other words, homomorphism should preserve immediate predecessors and the conflict relation.

@@

\begin{definition}[Category $\catname{EHyp_{\Sigma}}$]
Such e-hypergraphs form a category $\catname{EHyp_{\Sigma}}$ where morphisms are e-hypergraph homomorphisms defined above.
In particular, it has a coproduct which is a disjoint union of e-hypergraphs and a tensor unit which is an empty e-hypergraphs with empty relations which makes it into a monoidal category.
\end{definition}

\begin{remark}
    Similarly to $\catname{Hyp_{\Sigma}}$ pushouts in $\catname{EHyp_{\Sigma}}$ are computed on the underlying sets of nodes and edges.
    However, unlike $\catname{Hyp_{\Sigma}}$ $\catname{EHyp_{\Sigma}}$ is not \textit{adhesive} meaning that not all pushouts along monos exist.
    Consider an example span below in Figure~\ref{fig:pushout_fail}.
\begin{figure}[h!]
    \centering
    \[    
    \scalebox{0.5}{\tikzfig{combinatorial_semantics/pushout_fail_example}
    }
    \]
    \caption{Pushout failure}
    \label{fig:pushout_fail}
\end{figure}

    In order to build a commuting square which is a pushout the immediate predecessors of $f$ should be identified in the pushout because any element of an e-hypergraph has only one immediate predecessor.
    However, such identification is not possible as these predecessors have different types.
    Luckily, we do not need arbitrary pushouts and pushouts that we actually need do exist.
    \update{In general, in the pushout $A \xrightarrow{f} D \xleftarrow{g} B$, $D$ would inherit the relations from $A$ and $B$ when the relations are not mutually exclusive.}
\end{remark}

@@

\update{I will build below $\Ecospans$ (E for extended) instead of $ECsp(\mathcal{C})$ for arbitrary $\mathcal{C}$ with colimits to avoid talking about pushouts in general below, because we use set difference.}
\begin{definition}[Category $\Ecospans$]
    We will build the category of \textit{extended} cospans of e-hypergraphs as follows.
    Let the objects of $\Ecospans$ be only discrete e-hypergraphs (sets of edges and nodes are empty as well as both child and conflict relations) of $\catname{EHyp_{\Sigma}}$ and morphisms from $W \to Z$ be the cospans

    \[
    W \xrightarrow{f} X \xrightarrow{g} A \xleftarrow{h} Y \xleftarrow{k} Z
    \] 
    where $A \in obj(\catname{EHyp_{\Sigma}})$, $W, X, Y, Z$ are discrete and $f,g,h,k$ are monos.
    Since $f$ and $k$ are monos, $W$ and $Z$ are essentially subgraphs of $X$ and $Y$ and hence we will introduce the following notation which better depicts that such cospans are effectively e-hypergraphs with \textit{external} interfaces

    \[
    X_{ext} \xrightarrow{f_{ext}} X \xrightarrow{f} A \xleftarrow{g} Y \xleftarrow{g_{ext}} Y_{ext}
    \]

@@

    \[
    \scalebox{0.75}{
        \tikzfig{combinatorial_semantics/isomorphic_e_cospans}
    }
    \]

@@

\Aleksei{Redraw this diagram in tikzit because importing 'quiver' introduces error}

% \[
% \begin{tikzcd}
% 	& X & A & {~Y} \\
% 	{X_{ext}} &&&& {Y_{ext}} \\
% 	& {X'} & {A'} & {Y'}
% 	\arrow[curve={height=-12pt}, from=2-1, to=1-2]
% 	\arrow[from=1-2, to=1-3]
% 	\arrow[from=1-4, to=1-3]
% 	\arrow[curve={height=12pt}, from=2-5, to=1-4]
% 	\arrow[curve={height=-12pt}, from=2-5, to=3-4]
% 	\arrow[from=3-4, to=3-3]
% 	\arrow[from=3-2, to=3-3]
% 	\arrow[curve={height=12pt}, from=2-1, to=3-2]
% 	\arrow["\alpha"{description}, from=1-2, to=3-2]
% 	\arrow["\beta"{description}, from=1-3, to=3-3]
% 	\arrow["\gamma"{description}, from=1-4, to=3-4]
% \end{tikzcd}
% \]

@@

Composition of two morphisms $f : X_{ext} \to Y_{ext}$ and $g : Y_{ext} \to Z_{ext}$ is computed in two stages
First, $A +_{f_{ext};f, g_{ext};g} B$ is computed which is a pushout below

% https://q.uiver.app/#q=WzAsMTAsWzAsMCwiZXh0KFgpIl0sWzEsMSwiWCJdLFsyLDIsIkEiXSxbMywwLCJZIl0sWzQsMCwiZXh0KFkpIl0sWzUsMCwiWSciXSxbNiwyLCJCIl0sWzcsMSwiWiJdLFs4LDAsImV4dChaKSJdLFs0LDMsIl57XFx1bGNvcm5lcn1BK197Zl97ZXh0fTtmLGdfe2V4dH07Z31CIl0sWzAsMV0sWzEsMl0sWzMsMiwiZiJdLFs0LDMsImZfe2V4dH0iXSxbNCw1LCJnX3tleHR9IiwyXSxbNSw2LCJnIiwyXSxbNyw2XSxbOCw3XSxbMiw5XSxbNiw5XV0=

\[
\scalebox{0.9}
{\tikzfig{combinatorial_semantics/pushout_e_cospans}}
\]

Then, new feet for $A +_{f_{ext};f,g_{ext};g} B$ are computed as below

\[
X_{ext} \xrightarrow{f_{ext}} X \uplus (Y' \setminus Y_{ext}) \xrightarrow{h} A +_{f_{ext};f,g_{ext};g} B \xleftarrow{k} Z \uplus (Y \setminus Y_{ext}) \xleftarrow{g_{ext}} Z_{ext}
\]

since $X,Y,Z,Y', Y_{ext}$ are all discrete, $\uplus$ and $\setminus$ are standard set operations.

% \[
% \begin{tikzcd}
% Y_{ext} \arrow[d] & 0 \arrow[l] \arrow[d]    \\
% Y'          & Y' \setminus Y_{ext} \arrow[l]
% \end{tikzcd}
% \]

% and then the disjoint union is just a coproduct. 
Defining cospans up to isomorphism, associativity of coproducts and pushouts, and the presence of an id-cospan turns this construction into a valid category.

Similarly to~\ref{def:cspd} this category has a symmetric monoidal product given by a coproduct in $\catname{EHyp_{\Sigma}}$ which makes it into a SMC.
More precisely, below are cospans for tensor product and symmetry. Tensor unit is given by the unit in $\catname{EHyp_{\Sigma}}$.

% https://q.uiver.app/#q=WzAsMTcsWzAsMCwiZXh0KFgpIl0sWzEsMCwiWCJdLFsyLDAsIkEiXSxbMywwLCJZIl0sWzQsMCwiZXh0KFkpIl0sWzAsMiwiZXh0KFgnKSJdLFsxLDIsIlgnIl0sWzIsMiwiQiJdLFszLDIsIlknIl0sWzQsMiwiZXh0KFknKSJdLFsyLDEsIlxcb3RpbWVzIl0sWzUsMSwiOj0iXSxbNiwxLCJleHQoWCkrZXh0KFgnKSJdLFs4LDEsIlgrWCciXSxbMTAsMSwiQStCIl0sWzEyLDEsIlkrWSciXSxbMTQsMSwiZXh0KFkpK2V4dChZJykiXSxbMCwxLCJmX3tleHR9Il0sWzQsMywiZ197ZXh0fSIsMl0sWzMsMiwiZyIsMl0sWzEsMiwiZiJdLFs1LDYsImYnX3tleHR9Il0sWzYsNywiZiciXSxbOCw3LCJnJyIsMl0sWzksOCwiZydfe2V4dH0iLDJdLFsxMiwxMywiZl97ZXh0fStmJ197ZXh0fSJdLFsxMywxNCwiZitmJyJdLFsxNSwxNCwiZytnJyIsMl0sWzE2LDE1LCJnX3tleHR9ICsgZydfe2V4dH0iLDJdXQ==
\[
\adjustbox{scale=0.6, center}{
\begin{tikzcd}
	{X_{ext}} & X & A & Y & {Y_{ext}} \\
	&& \otimes &&& {:=} & {X_{ext}+ext(X')} && {X+X'} && {A+B} && {Y+Y'} && {Y_{ext}+Y'_{ext}} \\
	{ext(X')} & {X'} & B & {Y'} & {Y'_{ext}}
	\arrow["{f_{ext}}", from=1-1, to=1-2]
	\arrow["{g_{ext}}"', from=1-5, to=1-4]
	\arrow["g"', from=1-4, to=1-3]
	\arrow["f", from=1-2, to=1-3]
	\arrow["{f'_{ext}}", from=3-1, to=3-2]
	\arrow["{f'}", from=3-2, to=3-3]
	\arrow["{g'}"', from=3-4, to=3-3]
	\arrow["{g'_{ext}}"', from=3-5, to=3-4]
	\arrow["{f_{ext}+f'_{ext}}", from=2-7, to=2-9]
	\arrow["{f+f'}", from=2-9, to=2-11]
	\arrow["{g+g'}"', from=2-13, to=2-11]
	\arrow["{g_{ext} + g'_{ext}}"', from=2-15, to=2-13]
\end{tikzcd}
}
\]

@@

\begin{remark}
    When writing things like $Y' \setminus Y_{ext}$ we mean $Y' \setminus g(Y')$ and so on.
\end{remark}

\begin{remark}
An example of such a morphism in $\Ecospans$ can be found in Figure~\ref{fig:e-cospan}.
The feet of a cospan are depicted on top and at the bottom from the grey-shaded area.
These rectangles correspond to $X_{ext}$, $X$, $Y$, and $Y_{ext}$ respectively.
\end{remark}

\begin{figure}[h!]
    \centering
    \[
    \scalebox{0.5}{
    \tikzfig{combinatorial_semantics/e-hypergraph-example}
    }
    \]
    \caption{Morphism in $\Ecospans$}
    \label{fig:e-cospan}
\end{figure}
